Fusing Filters with Integer Linear Programming

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
ILP section
changed the order of fusion options to match with figure, added some
text

\begin{figure}[ht!]
\begin{center}
\includegraphics[scale=0.5]{figures/ex2-normalizeInc.pdf}
\end{center}
\caption{Possible clusterings for \texttt{normalizeInc}}
\end{figure}

The problem of finding the ``best'' clustering according to some criteria is known to be NP-hard~\cite{darte1999complexity}.
For example, in the special case of single-typed fusion, where all loops have the same iteration size, finding a clustering with the minimal number of loops can be done in polynomial time.
However, to find the minimal number of manifest arrays, each node may need to be considered in several clusters, and the problem becomes NP-hard.
Similarly, after adding multiple iteration sizes that cannot be fused together, the problem becomes NP-hard.

The clusterings of a given program are not unique; there are many valid clusterings, far too many to feasibly enumerate all of them.
There may even be multiple clusterings with the minimal number of loops. For example, consider the following program:

\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}

For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@). The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@). For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

Our objective, then, is not to simply reduce the number of loops as in~\cite{kennedy1993typedfusion}, but to reduce overall memory bandwidth, manifest arrays, and loops.
The objective of reducing manifest arrays, or maximising array contraction, has been discussed in~\cite{gao1993collective, darte2002contraction}.

In order to deal with the combinatorial explosion of potential clusterings, we describe an integer linear programming formulation.
Integer linear programming problems are defined as a set of variables, an objective linear function and a set of linear constraints.
The integer linear solver finds an assignment to the variables that minimises the objective function, while satisfying all constraints.
This problem is NP-hard, but existing ILP solvers are able to adequately solve practical problems.
Most combinator programs are small enough for this approach to be feasible, but for larger problems an approximation could be found, within a certain threshhold of the objective function.

@@

The dependency graph is translated to an integer linear program. The integer linear program has some integral, boolean and real variables, an objective function to minimise, and a set of constraints that the variables must conform to. Finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, but existing ILP solvers tend to be sufficient for realistic problem sizes. For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still gives the exact answer for small problems.

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used to ensure that the clustering is acyclic -- that is, there must be an ordering to execute each clusters' dependency clusters before executing the cluster itself.
For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$. This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor. If two nodes are to be fused together, their $\pi$ values must be the same.

As an example of a cyclic clustering, consider:
\begin{code}
cycle us
 = let xs  = map (+1) us          (C1)
       sum = fold xs              (C2)
       zs  = map (+sum) xs        (C1)
\end{code}
There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.

>>>>

\begin{figure}[ht!]
\begin{center}
\includegraphics[scale=0.5]{figures/ex2-normalizeInc.pdf}
\end{center}
\caption{Possible clusterings for \texttt{normalizeInc}}
\label{f:normInc}
\end{figure}

\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering. Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function. Compare with prior work on typed fusion that only tries to minimise the number of clusters.} \TODO{Also give a quick rundown of what ILP actually is.}
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A2) (B1)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}
% swapped A and B, so that the first option matches with the left graph, the second with the right graph in the diagram
For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}. The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@). For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@. For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

The @normalizeInc@ example demonstrates why it is, in general, not sufficient, to simply minimize the number of loops in a program: we have to assign weights to different types of loops and take those into account when choosing a possible clustering. Ordered typed fusion \cite{kennedy1993typedfusion} addresses this problem in the context of fusing imperative parallel and sequential loops, but also only in the absence of filter-like operations. Our problem can be directly mapped to an integer linear program with some integral, boolean and real variables, an objective function to minimise, and a set of contraints the variables must conform to. While finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, we can build on exisiting work here and use the best ILP-solver available, which tend to be sufficient for realistic problem sizes.  For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still results in a very efficient, if not necessarily optimal program.

%For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@). The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@). For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

@@

% The dependency graph is translated to an integer linear program. The integer linear program has some integral, boolean and real variables, an objective function to minimise, and a set of constraints that the variables must conform to. Finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, but existing ILP solvers tend to be sufficient for realistic problem sizes. For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still gives the exact answer for small problems.

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used to ensure that the clustering is acyclic. For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$. This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor. If two nodes are to be fused together, their $\pi$ values must be the same. \ben{I don't understand the ``acyclic'' part. If we don't add these constraints, then how might the clustering turn out to be ``cyclic''?}
